---
title: "The method argument"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The method argument}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>

img {
  border: 0px;
}

</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  dpi = 160,
  out.width = "80%",
  collapse = TRUE,
  comment = "#>"
)

```

Almost every function in **ggdensity** accepts a `method` argument---this is true for `geom_hdr()` and other layer functions
(`geom_hdr_lines()`, `geom_hdr_points()`, ...),
as well as `get_hdr()` and `get_hdr_1d()`.
This vignette summarizes the many ways in which the `method` argument can be specified; 
first looking at it from a more basic perspective,
then from the perspective of a developer wanting to implement additional estimators.

## Using **ggdensity**'s `method_*()` functions

First, let's load the necessary packages and generate some sample data.

```{r setup}
library("ggplot2"); theme_set(theme_minimal())
library("ggdensity")

theme_set(theme_minimal(8))
theme_update(
  # Suppressing legends for readability
  legend.position = "none"
)
```

```{r}
set.seed(1) 

df <- 
  data.frame(
    x = rnorm(500),
    y = rnorm(500)
  )

ggplot(df, aes(x, y)) +
  geom_point()
```

The easiest way to plot HDRs with `geom_hdr()` (or any other layer function from **ggdensity**) with a specified density estimator 
is to provide a character object to the `method` argument:
```{r, fig.show="hold", out.width="45%", fig.align = "default"}
ggplot(df, aes(x, y)) +
  geom_hdr(method = "kde")

ggplot(df, aes(x, y)) +
  geom_hdr(method = "mvnorm")

ggplot(df, aes(x, y)) +
  geom_hdr(method = "histogram")

ggplot(df, aes(x, y)) +
  geom_hdr(method = "freqpoly")
```

However, as of **ggdensity** v1.0.0 there is an alternative approach---providing a `method_*()` function call:
```{r, fig.show="hold", out.width="45%", fig.align = "default"}
ggplot(df, aes(x, y)) +
  geom_hdr(method = method_kde())

ggplot(df, aes(x, y)) +
  geom_hdr(method = method_mvnorm())

ggplot(df, aes(x, y)) +
  geom_hdr(method = method_histogram())

ggplot(df, aes(x, y)) +
  geom_hdr(method = method_freqpoly())
```

The default behaviors of these two approaches are the same and always will be---in this way, they are completely interchangeable. 
However, the `method_*()` function call is required to estimate HDRs with non-default estimator parameters. 
For example, we can set the `adjust` parameter to apply a multiplicative adjustment to the heuristically determined bandwidth in `method_kde()`:

```{r}
ggplot(df, aes(x, y)) +
  geom_hdr(method = method_kde(adjust = 1/2))
```

The relevant parameters for each method are documented in their respective `?method_*` help pages.
Note that these parameters can not be provided to `geom_hdr()` or `stat_hdr()` and thus are not accessible if 
a character value is provided to `method`.

The `method` argument of `get_hdr()` functions in the same way:

```{r}
res <- get_hdr(df, method = method_kde(adjust = 1/2))

str(res)
```

For details on the output of `get_hdr()`, see `?get_hdr`.


### `method_*_1d()` functions

In **ggdensity**, it is possible to estimate and plot 1-dimensional HDRs with `geom_hdr_rug()` and `get_hdr_1d()`.
These functions also accept a `method` argument, but they do not accept the previously discussed `method_*()` functions.
Instead they accept the 1-dimensional analogues: `method_*_1d()`.

```{r, fig.show="hold", out.width="45%", fig.align = "default"}
ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = method_kde_1d())

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = method_norm_1d())

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = method_histogram_1d())

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = method_freqpoly_1d())
```

Just like we saw with `geom_hdr()`, `geom_hdr_rug()` also accepts character values for `method`:

```{r, fig.show="hold", out.width="45%", fig.align = "default"}
ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = "kde")

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = "norm")

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = "histogram")

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = "freqpoly")
```

Because the return values of the `method_*()` functions are incompatible with the 1-dimensional
HDR estimation procedure, if a 2-dimensional method is specified the following error message is issued:

```{r, fig.show = "hide"}
ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hdr_rug(method = method_kde()) 
```

Lastly, we see that the `method` argument of `get_hdr_1d()` behaves similarly.

```{r}
res <- get_hdr_1d(df$x, method = method_kde_1d())

str(res)
```

Again, for details on the above output of `get_hdr_1d()`, see `?get_hdr_1d`.


## A detailed look at `method_*()` functions

Now that we understand the ways in which `method` can be specified
let's look at the internals of the `method_*()` functions.
Note: the implementations discussed in this section depend heavily on topics in functional programming,
especially 
[closures](https://adv-r.hadley.nz/environments.html?q=closures#function-environments) and
[function factories](https://adv-r.hadley.nz/function-factories.html).
While not necessary, a good understanding of these ideas is helpful---the 
linked chapters from Hadley Wickham's *Advanced R* are a great place to start.

Looking at the definition of `method_kde()`, we see that it is a function of `h` and `adjust`,
returning a closure with arguments `data`, `n`, `rangex`, and `rangey`.
The closure estimates passes the `x` and `y` columns of `data` to `MASS::kde2d()`, 
returning the estimated density evaluated on a grid with columns `x`, `y`, and `fhat`.
This closure is what `geom_hdr()` expects as its `method` argument,
and is how the HDRs are estimated (via `get_hdr()`).

```{r, collapse = TRUE, comment = ""}
method_kde
```

Both `method_histogram()` and `method_freqpoly()` behave similarly, 
accepting parameters governing the density estimation procedure and returning a closure with arguments
`data`, `n`, `rangex`, and `rangey`.
However, these functions are significantly more complicated as
the density estimation procedures are implemented entirely in **ggdensity**.

`method_mvnorm()` is different in a few ways.
The closure it returns is a function of just one argument: `data`.
This is because it does not return the estimated density evaluated on a grid.
Instead, it returns yet another closure with (vectorized) arguments `x` and `y`.
As in `method_kde()`, the return value of the closure is a representation of the estimated pdf.
The difference is the manner in which the pdf is represented.
Whereas before we had a pdf defined by a discrete approximation on a grid,
we now have an explicit definition of the pdf in terms of `x` and `y`.

```{r, collapse = TRUE, comment = ""}
method_mvnorm
```

To summarize each of the above cases:
in the first example, the `method_*()` function returned a closure with arguments 
`data`, `n`, `rangex`, and `rangey` which itself returned the estimated density evaluated on a grid;
in the second, the `method_*()` function returned a closure with a single argument, `data`, 
which itself returned a closure with arguments `x` and `y`, representing the estimated density explicitly.
In both cases, the `method_*()` function can have any number of parameters governing the density estimation procedure.

These are the two ways the `method` argument may be specified.
The first is necessary for cases in which an explicit definition of the estimated density is not computationally feasible 
(for example, KDEs).
The second is an easier option for the cases in which a closed form of the estimated density is available 
(typically parametric estimators fall into this category).

Let's look at how we might define our own `method_*()` functions in each case,
beginning with a simple parametric estimator.

### Implementing `method_mvnorm_ind()`

In **ggdensity**, `method_mvnorm()` estimates HDRs based on the parametric multivariate normal model.
If we wanted to fit a simpler model in which the data is further assumed to be independent,
we could implement `method_mvnorm_ind()`.

```{r}
method_mvnorm_ind <- function() {
  
  function(data) {
    
    xbar <- mean(data$x)
    ybar <- mean(data$y)
    
    sx <- sd(data$x)
    sy <- sd(data$y)
    
    # joint pdf is simply the product of the marginals
    function(x, y) dnorm(x, xbar, sx) * dnorm(y, ybar, sy)
  }
  
}
```

To use our `method_mvnorm_ind()`, we simply supply it to `geom_hdr()`'s `method` argument.
```{r}
ggplot(df, aes(x, y)) +
  geom_hdr(method = method_mvnorm_ind())
```

If we transform our data to have non-zero covariance
we still see the major and minor axes of the contours coincide exactly with the plot axes---exactly
what we would expect with this (incorrectly) constrained model.

```{r}
A <- matrix(c(
  2*cos(pi/6), -2*sin(pi/6),
  1*sin(pi/6),  1*cos(pi/6)
), byrow = TRUE, ncol = 2)

df_rot <- as.data.frame(as.matrix(df) %*% A)
colnames(df_rot) <- c("x", "y")

ggplot(df_rot, aes(x, y)) +
  geom_hdr(method = method_mvnorm_ind()) +
  geom_point(size = .4) +
  coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6))
```

Notice, `method_mvnorm_ind()` accepts no arguments.
The density estimation procedure is so simple that there are no parameters to govern it.
To allow for circular models in which the fitted variances are required to be equal,
we can implement a `circular` argument.
```{r}
method_mvnorm_ind <- function(circular = FALSE) {
  
  function(data) {
    
    xbar <- mean(data$x)
    ybar <- mean(data$y)
    
    if (circular) {
      sx <- sd(c(data$x, data$y))
      sy <- sx
    } else {
      sx <- sd(data$x)
      sy <- sd(data$y)
    }
    
    function(x, y) dnorm(x, xbar, sx) * dnorm(y, ybar, sy)
  }
  
}
```

Now, the contours are perfectly circular (almost).

```{r}
ggplot(df_rot, aes(x, y)) +
  geom_hdr(method = method_mvnorm_ind(circular = TRUE)) +
  geom_point(size = .4) +
  coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6))
```

In the above plot, the upper and lower portions of the HDRs are cut off.
This is because the default behavior of **ggdensity** is to not draw HDRs outside of the "bounding box" around observed data.
This is *not* because we are using a custom `method_*()` function.
To fix this, we need to either set a better `ylim` value for `geom_hdr()` or specify a larger range in `scale_y_continuous()`.

```{r, fig.show="hold", out.width="45%", fig.align = "default"}
ggplot(df_rot, aes(x, y)) +
  geom_hdr(method = method_mvnorm_ind(circular = TRUE), ylim = c(-6, 6)) +
  geom_point(size = .4) +
  coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6))

ggplot(df_rot, aes(x, y)) +
  geom_hdr(method = method_mvnorm_ind(circular = TRUE)) +
  geom_point(size = .4) +
  scale_y_continuous(limits = c(-6, 6)) +
  coord_fixed(xlim = c(-6, 6), ylim = c(-6, 6))
```

Notice, neither of these approaches involve arguments to `method_mvnorm_ind()`.
Internally, the closure returned by `method_mvnorm_ind()` is used by `get_hdr()`,
along with information from the `scales` associated with the `ggplot` object.
It is the `scales` that need adjusting, not anything related to the `method` argument.


### Implementing `method_random()`

To illustrate the other case, in which the object returned by the closure is the estimated density evaluated on a grid, we implement `method_random()`.
This "estimates" a density on a grid with random numbers from `runif()`
(this toy "density estimator" is useless; notice the `data` argument is not being used to determine `df_grid$fhat`).

```{r}
method_random <- function() {
  
  function(data, n, rangex, rangey) {
    
    df_grid <- expand.grid(
      x = seq(rangex[1], rangex[2], length.out = n),
      y = seq(rangey[1], rangey[2], length.out = n)
    )
    
    df_grid$fhat <- runif(n^2)
    
    df_grid
    
  }  
  
}

```

This simple example shows the general structure of `method_*()` functions which deal with gridded density estimates.
The returned closure has additional arguments `n`, `rangex`, and `rangey` which define the grid;
and the closure returns the grid as a `data.frame` with columns `x`, `y`, and `fhat`, where `fhat` is the (potentially unnormalized) density estimate.

As expected, the HDRs resulting from this method are just random noise.

```{r}
ggplot(df, aes(x, y)) +
  geom_hdr(method = method_random())
```

## The `method_*_1d()` functions

We saw before that **ggdensity** uses `method_*_1d()` functions for the estimation of 1-dimensional densities.
The internals of these functions are very similar to the `method_*()` functions,
the only differences are slight changes to the arguments and return values of the returned closures.

Looking at the definition of `method_kde_1d()`, we see the returned closure has arguments `x`, `n`, and `range`.
This is very similar to `method_kde()`, the only difference is we are now dealing with univariate data:
the vector argument `x` is used instead of `data`, and we have a single `range` parameter instead of `rangex` and `rangey`.
Similarly, the closure now returns the estimated density evaluated on a univariate grid, with columns `x` and `fhat` instead of the bivariate grid with columns `x`, `y`, and `fhat`.
Finally, see that `method_kde_1d()` accepts several arguments governing the density estimation procedure just like `method_kde()`.

```{r, collapse = TRUE, comment = ""}
method_kde_1d
```

Estimated univariate densities can also be represented explicitly, as illustrated by `method_norm_1d()`.
Comparing this to the previously discussed `method_mvnorm()` we see that little has changed:
the closure is now a function of a vector `x` instead of `data` and returns a function of one variable (`x`) instead of two (`x` and `y`).
```{r, collapse = TRUE, comment = ""}
method_norm_1d
```

Additional `method_*_1d()` functions can be implemented in the same way as the 2-dimensional `method_*()` functions,
so long as the returned closure is structured in one of the two ways we have seen here.





